---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# zarr

<!-- badges: start -->
<!-- badges: end -->

`zarr` is a package to create and access Zarr stores using native R code. This package implements a native R driver for `zarr` stores. It is designed against the specification for Zarr core version 3.

### Basic usage

The easiest way to zarrify your data is simply to call `as_zarr()` on your R array-like object.
```{r as}
library(zarr)

x <- array(runif(400), c(5, 20, 4))
z <- as_zarr(x)
z
```
`z` is an in-memory `zarr` object, with the R object broken up into chunks (if the dimensions are substantially large enough to warrant that) and compressed. If you prefer to persist your R object on file, provide a location where you want to store the data:
```{r persist}
fn <- tempfile(fileext = ".zarr")
z <- as_zarr(x, fn)
z
```
more
```
unlink(fn)
```
### Development

This package is in the early phases of development and should not be used for production environments. Things may fail and you are advised to ensure that you have backups of all data that you put in a Zarr store with this package.

Like Zarr itself, this package is modular and allows for additional stores, codes, transformers and extensions to be added to this basic implementation. If you have specific needs, open an [issue on Github](https://github.com/R-CF/zarr/issues) or, better yet, fork the code and submit code suggestions via a pull request. Specific guidance for developers is being drafted.

