---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# zarr

<!-- badges: start -->
[![R-CMD-check](https://github.com/R-CF/zarr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-CF/zarr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`zarr` is a package to create and access Zarr stores using native R code. It is designed against the specification for Zarr core version 3. Supported stores comprise memory, the file system and HTTP.

Zarr v.2 stores can be read and will be represented internally in the v.3 format. 

## Basic usage

The easiest way to zarrify your data is simply to call `as_zarr()` on your R vector, matrix or array.
```{r as}
library(zarr)

x <- array(1:400, c(5, 20, 4))
z <- as_zarr(x)
z
```
This uses information from the array (data type, dimensions) in combination with default Zarr settings (chunk size, data layout, compression). `z` is an in-memory `zarr` object consisting of a single array, with the R object broken up into chunks (if the dimensions are substantially large enough to warrant that) and compressed.

If you prefer to persist your R object to file, provide a location where you want to store the data. It is recommended that the name of the store uses the ".zarr" extension. At the same time you may give your Zarr array a name that you can later use to retrieve the array and its data.
```{r persist}
fn <- tempfile(fileext = ".zarr")
z <- as_zarr(x, name = "my_array", location = fn)
z
```
The total size that is printed to the console when inspecting a Zarr object sums up the size of both types of files in the root directory and any sub-directories.

##### Reading and writing Zarr arrays
You can access the data in a Zarr object through its arrays. If you didn't specify a name when using function `as_zarr()`, the Zarr object can only hold a single array, which is located at the root `"/"` of the Zarr object. If you did specify a name you should indicate the full path starting with a slash `"/"` for the root:
```{r array}
# Get the array using list-like access on the Zarr object
arr <- z[["/my_array"]]
arr

# Index the Zarr array like a regular R array
# Indexes are 1-based, as usual in R (Zarr specifies everything as 0-based)
arr[1:2, 11:16, 3]
```
You can also write to a Zarr array, but the process is a bit more complicated (due to a quirk in R):
```{r write}
arr$write(-99L, selection = list(2:3, 5:7, 1))
arr$write(NA_integer_, selection = list(1:5, 1, 1))
arr[, 1:10, 1]
```
The data in the Zarr array is of type "int32", the standard R integer. When writing data you should make sure that the object to be written is of the correct type, so using "-99L" and "NA_integer_" here. Numeric data is stored by default as "float64", logical data as "int8".
```{r, include = FALSE}
unlink(fn)
```
## Installation
Installation from CRAN of the latest release:
```
install.packages("zarr")
```
You can install the development version of `zarr` from [GitHub](https://github.com/R-CF/zarr) with:
```
# install.packages("devtools")
devtools::install_github("R-CF/zarr")
```
## Development
This package is in the early phases of development and should not be used for production environments. Things may fail and you are advised to ensure that you have backups of all data that you put in a Zarr store with this package.

Like Zarr itself, this package is modular and allows for additional stores, codes, transformers and extensions to be added to this basic implementation. If you have specific needs, open an [issue on Github](https://github.com/R-CF/zarr/issues) or, better yet, fork the code and submit code suggestions via a pull request. Specific guidance for developers is being drafted.
