---
title: "3. Working with attributes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
Zarr groups and arrays can have attributes associated with them. Attributes are typically character strings describing some notable feature of a group or array, such as the physical unit of the data in an array or the intended use of one or more arrays. Attributes may also be numeric, such as in the case of a range of valid values for an array. Most (simple) R objects can be used in attributes but it is not recommended to use compound (e.g. `list()`) or large objects. Matrices and arrays should be avoided where possible because there is no good way to indicate the storage order of the values. Attributes are best held small for performance reasons. Attributes are stored in the same "zarr.json" metadata files as the structural information on groups and arrays and large attributes will thus slow down the opening and writing of groups and arrays.

Attribute management is very easy, using two methods `set_attribute()` and `delete_attributes()`, available for both `zarr_group` and `zarr_array` instances.

```{r setup}
library(zarr)

# Create an in-memory Zarr object
z <- create_zarr()

# Set attributes in the root group, a.k.a. global attributes
z[["/"]]$set_attribute("title", "Data set for intelligent analysis of foo.")
z[["/"]]$set_attribute("creator", "ACME Inc.")
z[["/"]]$set_attribute("license", "free (as in free lunch)")
z

# Create an array and attach some attributes
x <- array(runif(20000), c(100, 200))
arr <- as_zarr(x, "bar", z[["/"]])
arr$set_attribute("title", "Bar for foo")
arr$set_attribute("valid_range", c(0, 1))
arr$set_attribute("actual_range", range(x))
arr 
```
You can programmatically retrieve the full set of attributes with the `attributes` field of `zarr_group` and `zarr_array`. The attributes are returned to the caller as a `list`.

You can update or overwrite an existing attribute simply by setting the new value for the attribute. In combination with accessing all attributes this allows you to update an existing attribute value too.
```{r attr_update}
arr$set_attribute("title", paste(arr$attributes[["title"]], "and baz too"))
arr
```
If you want to delete one or more attributes, use the `delete_attributes()` method which takes a vector of attribute names to delete.
```{r attr_delete}
arr$delete_attributes(c("valid_range", "actual_range"))
arr
```
The `(*)` after the the heading "Attributes:" indicates that there are unsaved changes to the attributes, meaning that the changes have not been persisted to a Zarr store. If you wish to persist those changes, simply call `arr$save()`.
