---
title: "4. HTTP stores"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(zarr)
```
The `zarr` package can read online stores over HTTP, meaning that Zarr stores published on a regular web site are accessible. Access is read-only. Unfortunately, there is no uniform mechanism to identify what Zarr arrays are published and where they are located on the web site. This package supports three mechanisms to access HTTP stores.

# Accessing HTTP stores

## Zarr v.2: Consolidated metadata
Zarr v.2 supports consolidated metadata, a JSON file called `.zmetadata` in the root of the HTPP store which lists all the groups and arrays present in the store. When a `.zmetadata` file is present, it will be automatically read and groups and arrays will be added to the `zarr` object to represent the contents of the HTTP store.

```{r edh}
# Publicly accessible Zarr v.2 HTTP store with consolidated metadata
z <- open_zarr("https://data.earthdatahub.destine.eu/public/test-dataset-v0.zarr")
z$hierarchy()
```

## Single-array store
If there is no consolidated metadata, the HTTP store will be queried for a `zarr.json` (v.3) or `.zarray` (v.2) JSON file, containing the description of a single Zarr array. If found, the array will added to the `zarr` object. The array may have attributes that describe the properties of the array, such as coordinates of the axes. These attributes have to be examined manually; the Zarr specification has no standards for attribute contents.

```{r afr}
z <- open_zarr("https://raw.githubusercontent.com/R-CF/zarr/main/inst/extdata/africa.zarr/tas")
z[["/"]]
```

The downside of this approach is that you need to have knowledge of the Zarr resource to be able to properly interpret the data.

## Group
When there is no consolidated metadata or a single array at the root of the HTTP store, a group is expected as the final option. The group should have attributes that explain what is contained in the store.

The below example uses sample data from the [Open Microscopy Environment](https://www.openmicroscopy.org) (OME) who have published the [OME-Zarr specification](https://ngff.openmicroscopy.org/specifications/index.html) for Zarr stores. You need to understand that specification to interpret the data.

```{r ome}
# A publicly accessible OME-Zarr data set
(z <- open_zarr("https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0044A/4007801.zarr"))

# The attribute "multiscales" indicates the details of the data sets in the store
# The "path" element is the path to the individual arrays
jsonlite::prettify(jsonlite::toJSON(z$root$attributes[["multiscales"]], auto_unbox = TRUE))

# Open the array at path "3"
z3 <- open_zarr("https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0044A/4007801.zarr/3")
z3[["/"]]
```

# Authentication
This package does not manage authentication for web resources. Authentication comes in many different forms, including supplier-specific protocols, and it is impossible to cater to them all. More basic forms of authentication may use HTTP session tokens that are inserted in the URL. If that is the case, you should manually get a session token and then insert it into the URL that is used to open the store in `zarr`.

For more complicated cases, it may be useful to write a small package on top of `zarr` to manage authentication. We'd love to hear of your efforts!

# Working with array data from a HTTP store
You can work with array data just like you would be array data from a file system store: index it like any other R array. Keep in mind, though, that the data will be fetched over your internet connection so be very judicious in your indexing. Downloading a full array is almost always a bad idea as Zarr arrays tend to be large. In the OME-Zarr example above, there are `532 * 2 * 988 * 256 * 271 = 72930271232` grid points, which expands to about 271GB of integer data in R!

A more intelligent way to download data is to look at the chunking of the array data. From the "axes" array in the "multiscales" attribute we can see that a chunk (the unit of downloading data) is all of the "x" and "y" extent for a single "[z, c, t]" tuple. The most efficient way of downloading the data is then to follow the chunking scheme and download one or a few of the "z", "c" and "t" values and all of "x" and "y".
